Using data.table to create data tables just like data frames

**Reminder option + command + i inserts chunk**

```{r}
library(data.table)
DF = data.frame(x=rnorm(9), y=rep(c("a","b","c"), each =3), z=rnorm(9))
head(DF,3)
```

Now to create the same table using data.table

```{r}
DT = data.table(x=rnorm(9), y=rep(c("a","b","c"),each=3),z=rnorm(9))
head(DT,3)
```

To see all the data tables in memory use the tables() command

```{r}
tables()
```

One difference between subsetting in data.table and data.frame...


```{r}
DT[c(2,3)]
## This will pull rows as opposed to columns like data.frame would
```

Calculating values for variables with expression:

With DT you can pass a list of functions you want to perform, where the functions are applied to variables named by columns

```{r}
DT[,list(mean(x), sum(z))]
```

To add new columns using data.table:

```{r}
DT[,w:=z^2]
## to add column w, use command := with values of Z square
```

A warning about adding new columns:

```{r}
DT2 <- DT
DT[, y:=2]

## DT2 is not a copy, and therefore future changes will impact the data of the orignal. The code above set column y = 2, which will be reflected in both DT and DTs; must use the copy function to create new DTs. 
```

Multiple operations:

```{r}
DT[,m:= {tmp <- (x+z); log2(tmp+5)}]

## multiple step functions to create new variable. Statements enclosed in curly brackets - tmp (temporary variable), adds x+z which is then passed to second statement.
```

Plyr like operations:

```{r}
DT[,a:= x>0]

## Adding variable A to the table, where x>0 will return a boolean value. If you want to summarize a variable by the boolean in a

DT[,b:= mean(x+w), by=a]

## This takes the mean of X+W when A is equal to true, and place that mean value in all the rows where A is euqal to true. Then, it will take the mean of X+W where A is equal to false and place that value into b where a is false.
```

Special Variables:

.N - An integer, length 1, containing the number

```{r}
set.seed(123);

DT <- data.table(x=sample(letters[1:3], 1E5, TRUE))
DT[, .N, by=x]

##.N is an integer linked one and it's a containment number of times that a particular group appears and so if I data table that has a large number of As, Bs, and Cs in it, so about 100,000 A B and Cs. To count the number of times each of those letters appear I can use data table, dot N, and then grouped by the X variable. What that will do is it will count, dot N is just count the number of times grouped by the X variable. It does that very fast as opposed to the equivalent operation, which is just doing that table of DT dollar sign X
```

Keys: A unique aspect of data tables is that they have keys, so if you set the key, it's possible to subset and sort a data table much more rapidly than you would be able to do with a data frame. Here, I'm going to create a data table, and it's going to have a variable X. And it's gonna have a variable Y and I'm gonna set the key for the data table to be the variable X. Then if I want a subset on the basis of x or if I put in quoted a here, it knows to go and look in the key, and the key is x, and it very quickly subsets the data to only the values of x that are equal to a. 

** THIS MIGHT BE USEFUL FOR PENDLETON'S DATA TO USE RANGES AS A KEY **

```{r}
DT <- data.table(x=rep(c("a","b","c"), each = 100), y=rnorm(300))
setkey(DT,x) 
DT['a']

```


Joins: You can also use keys to facilitate joints between data tables so for example here I've created two data tables where they have a variable X and a variable Y and in this case the second data table has a variable Z. I can set the key in both cases to be equal to x, so the same key for both data tables. And then I can merge them together. This is actually quite a bit faster than merging with the data frame as long as you have the same key for both data tables. It can be very fast. 

```{r}
DT1 <- data.table(x=c('a','a','b','dt1'), y=1:4)
DT2 <- data.table(x=c('a','b','dt2'), z=5:7)
setkey(DT1, x); setkey(DT2,x)
merge(DT1, DT2)
```
 
Data tables read faster from the disk.  
 
 
```{r}
big_df <- data.frame(x=rnorm(1E6), y=rnorm(1E6))
file <- tempfile()
write.table(big_df, file=file, row.names=FALSE, col.names=TRUE, sep="\t", quote=FALSE)
system.time(fread(file))
```

```{r}
system.time(read.table(file, header=TRUE, sep="\t"))
```

