Editing Text Variables

```{r}
if(!file.exists("./data")){dir.create("./data")}
fileUrl <- "https://data.baltimorecity.gov/api/views/dz54-2aru/rows.csv?accessType=DOWNLOAD"
download.file(fileUrl, destfile = "./data/cameras.csv", method="curl")
cameraData <- read.csv("./data/cameras.csv")
names(cameraData)
```

To simplify interactive querying and scripting, set all columns names to lower case:

```{r}
tolower(names(cameraData))
```

If you want to remove periods added by R upon import, utilize the strsplit() function

```{r}
splitNames = strsplit(names(cameraData),"\\.")

splitNames[[5]]
```

Looking at index 6, the original string has been split:

```{r}
splitNames[[6]]
```

Want to keep the first element of the list, so by using the sapply() function:
 
```{r}
firstElement <- function(x){x[1]}
sapply(splitNames, firstElement )
```


Editting text variabels with Peer Review data:

```{r}
fileUrl1 <- "https://dl.dropboxusercontent.com/u/7710864/data/reviews-apr29.csv"
fileUrl2 <- "https://dl.dropboxusercontent.com/u/7710864/data/solutions-apr29.csv"
download.file(fileUrl1, destfile = "./data/reviews.csv", method="curl")
download.file(fileUrl2, destfile = "./data/solutions.csv", method="curl")
reviews <- read.csv("./data/reviews.csv"); 
solutions <- read.csv("./data/solutions.csv")
head(reviews, 2)
```
```{r}
head(solutions, 2)
```

Checking for names and utilizing the sub() function to remove underscores

```{r}
names(reviews)
```
```{r}
sub("_", "", names(reviews))
```

If in the situation where you need to replace multiple characters in a string (intermittant underscores for example), utilize the gsub function as the sub() function will only replace the first instance in a string

```{r}
testName <- "this_is_a_test"
sub("_", "", testName)
```
Now with gsub()
```{r}
gsub("_", "", testName)
```

Searching for specific values or variable names:

Find all the intersections that include "Alameda" as one of the rows using the grep() function - will take as input a search string in all of a given variable and return the element positions 

```{r}
grep("Alameda", cameraData$intersection)
```


```{r}
table(grepl("Alameda", cameraData$intersection))
```

Using grepl() to subset data - suppose you want to find all the cases where alameda does not appear; use this to subset data for only certain cases you want to find

```{r}
cameraData2 <- cameraData[!grepl("Alameda", cameraData$intersection),]
cameraData2
```

Using the grep() function, you can set the value=TRUE to return the string as opposed to the position of the string in the data set

```{r}
grep("Alameda", cameraData$intersection, value=TRUE)
```

Another thing you can do with the grep() function is look for values that do not appear which return integer(0) if it doesn't appear.
```{r}
grep("JeffStreet", cameraData$intersection)
```

If you need 0 value return, use the length() function to return value.

```{r}
length(grep("JeffStreet", cameraData$intersection))
```

MORE USEFUL STRING FUNCTIONS

Finding number of characters in a string:
```{r}
library(stringr)
nchar("Jeffrey Leek")
```
Taking a substring of a string
```{r}
substr("Jeffrey Leek", 1, 7)
```

Paste two strings together:
```{r}
paste("Jeffrey", "Leek")
```

To paste strings together with no paste utilize the paste0() function

```{r}
paste0("Jeffery","Leek")
```

To trim white space off of strings:
```{r}
library(stringr)
str_trim("Jeff         ")
```

