Simulating data:

```{r}
x <- rnorm(100)
hist(x) #default y axis for hist() is frequency
```

Setting second variable Y and plotting against X:

```{r}
y <- rnorm(100)
plot(x,y)
```

To adjust the margins, you set a vector of four integers to argument mar =. Order of margins is south, east, north, west.

```{r}
par(mar = c(2,2,2,2))
plot(x,y) #you can lose margins if you push margins too much
```

To change the plotting symbol, use the pch argument:

```{r}
plot(x,y, pch=11)
```

To add title and text:

```{r}
plot(x,y, pch=7, col = "red")
title("Scatterplot")
text(-2, -2, "Label") #sets text at the coordinates
legend("topleft", legend = "Data", pch =7, col = "red")
fit <- lm(y~x)
abline(fit, lwd=3, col = "blue")



```

If you want two plots with same relationship to a given variable:

```{r}
z <- rpois(100,2)
par(mfrow = c(2,1))
par(mar = c(2,5,2,5))
plot(x,y, pch=7, xlab = "weight", ylab = "height", main = "Scatterplot", col = "red")
plot(x,z, pch=8, xlab = "weight", ylab = "height", main = "Scatterplot", col = "red")
legend("topright", legend = "Data", pch =7, col = "red")
fit <- lm(y~x)
abline(fit, lwd=3, col = "blue")

```

To create a scatter plot where points differentiate between factor levels by color

```{r}
x <- rnorm(100)
y <- x + rnorm(100)
g <- gl(2,50, labels = c("Male", "Female"))

plot(x,y)
points(x[g == "Male"], y[g == "Male"], col = "green")
points(x[g == "Female"], y[g == "Female"], col = "blue")

```

