Dimension Reduction:

Principle Components Analysis and Singular Value Decomposition are important techniques in the exploratory data anlysis phase and also in formal modelling. 

Starting with some matrix data...

```{r}
set.seed(12345)
par(mar = rep(0.2, 4))
dataMatrix <- matrix (rnorm(400), nrow = 40)
image(1:10, 1:40, t(dataMatrix)[, nrow(dataMatrix):1])
```

Image is noisy, so applying a hierarchical cluster on it:

```{r}
par(mar = rep(0.2, 4))
heatmap(dataMatrix)
```

Still, not real pattern emerges from the data, SO what if a pattern is added to data? You can loop through all the rows, and on a random row flip a coin, and if it turns out to be a one I just add a pattern, in which five of the columns have a mean of zero and the other five have a mean of three. So, you of add a little shift here across the columns.

```{r}
set.seed(678910)
for (i in 1:40){
        #flip a coin
        coinFlip <- rbinom(1, size = 1, prob = 0.5)
        # if coin is heads, add a common pattern to that row
        if(coinFlip){
                dataMatrix[i,] <- dataMatrix[i,] + rep(c(0,3), each = 5)
        }
}
```

So now if I plot the data you see that their at the right, on the right hand five columns are a little bit more yellow which means that they have a higher value and in the left hand side five columns which are, which are little bit more red which means they've a lower value. That's because some of the rows have a mean of three in the right hand side 

```{r}
heatmap(dataMatrix)
```

To take a closer look at the patterns in rows/columns to look at marginal means

```{r}
hh <- hclust(dist(dataMatrix))
dataMatrixOrdered <- dataMatrix[hh$order, ]
par(mfrow = c(1,3))
image(t(dataMatrixOrdered)[,nrow(dataMatrixOrdered):1])
plot(rowMeans(dataMatrixOrdered), 40:1, , xlab = "Row Mean", ylab = "Row", pch = 19)
plot(colMeans(dataMatrixOrdered), xlab = "Column", ylab = "column mean", pch = 19)

```

Components of the SVD - u and v:

```{r}
svd1 <- svd(scale(dataMatrixOrdered))
par(mfrow = c(1,3))
image(t(dataMatrixOrdered)[,nrow(dataMatrixOrdered):1])
plot(svd1$u[,1], , xlab = "Row", ylab = "First left singular vector", pch = 19)
plot(svd1$v[,1], xlab = "Column", ylab = "First right singular vector", pch = 19)

```

Components of the SVD = Variance Explained

```{r}
par(mfrow = c(1,2))
plot(svd1$d, xlab = "Column", ylab = "Singular value", pch = 19)
plot(svd1$d^2/sum(svd1$d^2), xlab = "Column", ylab = "Prop. of variance explained", 
     pch = 19)
```

Relationship to principal componenets

```{r}
svd1 <- svd(scale(dataMatrixOrdered))
pca1 <- prcomp(dataMatrixOrdered, scale = TRUE)
plot(pca1$rotation[,1],svd1$v[,1], pch = 19, xlab = "principal Component 1",
     ylab = "Right Singular Vector 1")
abline(c(0,1))
```

Components of the SVD - variance explained

```{r}
constantMatrix <- dataMatrixOrdered*0
for(i in 1:dim(dataMatrixOrdered)[1]){constantMatrix[i,] <- rep(c(0,1), each=5)}
svd1 <- svd(constantMatrix)
par(mfrow=c(1,3))
image(t(constantMatrix)[,nrow(constantMatrix):1])
plot(svd1$d, xlab="Column", ylab="Singluar value", pch=19)
plot(svd1$d^2/sum(svd1$d^2), xlab = "Column", ylab = "Prop. of variance explained", 
     pch =19)
```

```{r}
set.seed(678900)
for(i in 1:40){
        #flip a coin
        coinflip1 <- rbinom(1, size = 1, prob = 0.5)
        coinflip2 <- rbinom(1, size = 1, prob = 0.5)
        #if coin is heads add a common pattern to that row
        if (coinflip1){
                dataMatrix[i,] <- dataMatrix[i,] + rep(c(0,5), each = 5)
        }
        if (coinflip2){
                dataMatrix[i,] <- dataMatrix[i,] + rep(c(0,5),5)
        }
}

hh <- hclust(dist(dataMatrix))
dataMatrixOrdered <- dataMatrix[hh$order, ]

svd2 <- svd(scale(dataMatrixOrdered))
par(mfrow = c(1,3))
image(t(dataMatrixOrdered)[, nrow(dataMatrixOrdered):1])
plot(rep(c(0,1), each = 5), pch = 19, xlab = "Column", ylab = "Pattern 1")
plot(rep(c(0,1), 5), pch = 19, xlab = "column", ylab = "Pattern 2")
```

```{r}
svd2 <- svd(scale(dataMatrixOrdered))
par(mfrow = c(1,3))
image(t(dataMatrixOrdered)[,nrow(dataMatrixOrdered):1])
plot(svd2$v[,1], pch = 19, xlab = "Column", ylab = "First right singular vector")
plot(svd2$v[,2], pch = 19, xlab = "Column", ylab = "Second right singular vector")
```

```{r}
svd1 <- svd(scale(dataMatrixOrdered))
par(mfrow = c(1,2))
plot(svd1$d, xlab = "Column", ylab = "Singular value", pch = 19)
plot(svd1$d^2/sum(svd1$d^2), xlab = "Column", ylab = "Percent of variance explained",
     pch = 19)
```

Missing Values:

```{r}
dataMatrix2 <- dataMatrixOrdered
# Randomly insert missing data
dataMatrix2[sample(1:100, size = 40, replace = FALSE)] <- NA
svd1 <- svd(scale(dataMatrix2)) ##Doesn't work!
```

```{r}
library(impute) 
dataMatrix2 <- dataMatrixOrdered
dataMatrix2[sample(1:100, size = 40, replace = FALSE)] <- NA
dataMatrix2 <- impute.knn(dataMatrix2)$data
svd1 <- svd(scale(dataMatrixOrdered)); svd2 <- svd(scale(dataMatrix2))
par(mfrow=c(1,2)); plot(svd1$v[,1],pch=19); plot(svd2$v[,1],pch=19)
```

