The Lattice Plotting System

Lattice functions:

xyplot(y~x | f * g, data)
- uses formula notation, hence the tildy 
- on left of tildy is y axis variable, right is x axis
- f and g are conditioning variables which are optional; indicate an interaction between the two variables
- second argument is data frame or list for function; if none passed, parent frame in the workspace is used

```{r}
library(lattice)
library(datasets)

xyplot(Ozone ~ Wind, data = airquality)
```

Slightly more complicated plot. Converted month into a factor, plotting ozone against wind by month.

```{r}
airquality <- transform(airquality, Month = factor(Month))
xyplot(Ozone ~ Wind | Month, data = airquality, layout = c(5,1))
```

Lattice behavior: lattice functions behave differently from base graphics
- base graphics plot data directly to graphics device; lattice grahpics return an object of class trellis
- print methods for lattice functions actually do the work of plotting the data on the graphcis device
- lattice functions return "plot objects" that can in principle be stored (usually better to svae the code + data)
- On the command line, trellis objects are auto-printed so that it appears the function is plotting the data
```{r}
p <- xyplot(Ozone ~ Wind, data = airquality) #If run, nothing will appear
```
```{r}
print(p) #explicit printing generates plot
```

Lattice Panel Functions
- Lattice functions have a panel function which controls what happens inside each panel of the plot
- The lattice package comes with deault panel functions, but you can supply your own if you want to customize what happens in each panel
- Panel functions receive the xy coordinates of teh data points in their panel (along with any optional arguments)

```{r}
set.seed(10)
x <- rnorm(100)
f <- rep(0:1, each = 50)
y <- x + f - f * x + rnorm(100, sd = 0.5)
f <- factor(f, labels = c("Group 1", "Group 2"))
xyplot(y ~ x | f, layout = c(2,1)) #Plot with 2 panels
```

To call custom panel:

```{r}
xyplot(y~x | f, panel = function(x,y,...){
        panel.xyplot(x,y,...) #first call the default panel function for 'xyplot'
        panel.abline(h = median(y), lty= 2) #add horizontal line at the median
        
})
```

Custom regression line:

```{r}
xyplot(y~x | f, panel = function(x,y,...){
        panel.xyplot(x,y,...) #first call the default panel function for 'xyplot'
        panel.lmline(x,y,col =2) #overlay a simple linear regression line

})
```

Many panel lattice plot:

```{r}

```

