Plotting systems in R

Base R plotting:

```{r}
library(datasets)
data(cars)
with(cars, plot(speed, dist))
```

Lattice Plotting:

```{r}
library(lattice)
state <- data.frame(state.x77, region = state.region)
xyplot(Life.Exp ~ Income | region, data = state, layout = c(4,1))
```

ggplot2 plotting:

```{r}
library(ggplot2)
data(mpg)
qplot(displ, hwy, data = mpg)
```


MORE ON BASE PLOTTING - Initializing graphics

Histograms:

```{r}
library(datasets)
hist(airquality$Ozone)
```

Scatter Plots:
```{r}
library(datasets)
with(airquality, plot(Wind, Ozone))
```

Boxplot
```{r}
library(datasets)
airquality <- transform(airquality, Month = factor(Month))
boxplot(Ozone ~ Month, airquality, xlab = "Month", ylab = "Ozone (ppb)")
```

Base plot with annotation:

```{r}
library(datasets)
with(airquality, plot(Wind, Ozone))
title(main = "Ozone and Wind in New York City")
```

Plotting same plot, but using main = for the title, and making all points from month of may blue:

```{r}
with(airquality, plot(Wind, Ozone, main = "Ozone and Wind in New York City"))
with(subset(airquality, Month ==5), points(Wind, Ozone, col="blue"))
```

Same plot with annotation:

type = "n" - sets up plots and initializes graphics device but does not actually plot anything

```{r}
with(airquality, plot(Wind, Ozone, main = "Ozone and Wind in New York City", type = "n"))
with(subset(airquality, Month == 5), points(Wind, Ozone, col = "blue"))
with(subset(airquality, Month != 5), points(Wind, Ozone, col = "red"))
legend("topright", pch = 1, col = c("blue", "red"), legend = c("May", "Other Months"))
```

Base plot with Regression Line:

```{r}
with(airquality, plot(Wind, Ozone, main = "Ozone and Wind in New York City", pch = 20))
model <- lm(Ozone ~ Wind, airquality)
abline(model, lwd = 2)
```

Multiple Base Plots:

Previous plots have all been single plots; if you want multiple plots on single device, you set the par(mfrow = c(1,2)) - one row, two columns - to get them side by side. Everytime plot is called, a new plot with be generated, one on left one on right.

```{r}
par(mfrow = c(1,2))
with(airquality, {
        plot(Wind, Ozone, main = "Ozone and Wind")
        plot(Solar.R, Ozone, main = "Ozone and Solar Radiation")
})
```

Multiple plots expanded: 

```{r}
par(mfrow = c(1,3), mar = c(4,4,2,1), oma = c(0,0,2,0))
with(airquality,{
        plot(Wind, Ozone, main = "Ozone and Wind")
        plot(Solar.R, Ozone, main = "Ozone and Solar Radiation")
        plot(Temp, Ozone, main = "Ozone and Temperature")
        mtext("Ozone and Weather in New York City", outer = TRUE)
})
```

